---
title: 'REPS Training: R and Excel Macros'
author: "Darren Kwong"
date: "July 20, 2017"
output: html_document
---

#Introduction
In this training, we will review how to prepare data tables in R, export them to an Excel file, and then format those newly created Excel tables. This lesson builds on previous R trainings, so this training assumes that the reader has a basic knowledge of [reading/writing data with R](J:/Research/!Projects/TRAIN -- REPS Training and PD/R/Lesson 01 Reading & Writing Data), [data wrangling with DPLYR](J:/Research/!Projects/TRAIN -- REPS Training and PD/R/Lesson 02 Data Wrangling with DPLYR), and [automating processes](J:/Research/!Projects/TRAIN -- REPS Training and PD/R/Lesson 03 Automating Data Processes).

For an interactive session, feel free to run the code below. Make sure to run the code by chunks. If you do that, then you don't have to worry about writing or saving over existing files. The code below has been developed, so that your files will be named dynamically.

#Set up
As always, our first step is to import relevant libraries and the sample data. For this exercise, we need:

-   *plyr* and *dplyr* for data manipulation;
-   *xlsx* to read and write Excel files;
-   *svDialogs* for soft coding; and
-   *RDCOMClient* for running macros in Excel.

```{r Import Libraries, warning = FALSE, message=FALSE, eval=FALSE}
library(plyr)
library(dplyr)
library(xlsx)
library(svDialogs)
library(RDCOMClient)
```
I created a sample data file for this demonstration. The file is saved in this Lesson's TRAIN folder within the "Sample Data" subfolder.
```{r Import Data, eval=FALSE}
sample_data <- read.xlsx("R:\\!Projects\\TRAIN -- REPS Training and PD\\R\\Lesson 05 Excel Macros\\Sample Data\\r_train_excel_macro_sample_data_20170720.xlsx", sheetIndex = 1)
```

#Create summary table
For the purposes of this exercise, let's say that we need to run a summary table that provides our program with information about their students per cohort. They want to know: (1) the number of students; (2) the average GPA; and (3) the graduation rate. We will present this information to them by cohort and the grand totals. 
```{r Summary Table, eval=FALSE}
summary_table <- sample_data %>% 
  mutate(Cohort = paste0(Semester," ",Year),
         grad_dummy = if_else(GRAD=="Yes", 1, 0)) %>% 
  group_by(Cohort) %>% 
  summarise(count = n(),
            Average.GPA = mean(mean_GPA),
            Grad.Rate = sum(grad_dummy)/count) %>% 
  ungroup()

summary_all <- summary_table %>% 
  summarise(Cohort = "All cohorts",
            count = sum(count),
            Average.GPA = mean(Average.GPA),
            Grad.Rate = sum(Grad.Rate)/count)

summary_table <- summary_table %>% 
  bind_rows(summary_all)
```

#Export to Excel and run macro
###The problem
After creating our summary table, we are ready to export it to Excel. In most situations, we would export the data as an ".xlsx" file. However, ".xlsx" files do not have embedded macros in them. And unfortunately, Excel requires that macros be embedded in the Excel document in order for them to run. In other words, we cannot run Excel macros based on VBA code outside of the Excel file.

###The workaround
I figured a workaround to this issue. To automate the process, we should:

1.  Record our macros in an Excel file (or if you're feeling more adventurus, you can type up the VBA from scratch)
2.  Save the Excel file in ".xlsm" format as a blank workbook
3.  Make a copy of that workbook and use it as a template for future data processes

###Recording macros in Excel
Recording macros in Excel is easy! Before your begin, list in order what you plan to record. When you are ready, go to the "Developer" Tab and click "Record Macro". Follow directions. When you are done, hit "Stop Recording". 

While recording the macro, here are a couple of important hot keys that I found useful:

- Ctrl + (down/up) arrow key = Go to last column
- Ctrl + (left/right) arrow key = Go to last row
- Ctrl + End = Last cell with data

###Running macro
In this scenario, I have already created an Excel workbook with a recorded macro. That file is saved here: [R:/!Projects/TRAIN -- REPS Training and PD/R/Lesson 05 Excel Macros/Sample VBA Macro](R:/!Projects/TRAIN -- REPS Training and PD/R/Lesson 05 Excel Macros/Sample VBA Macro). Feel free to check out the script via the .vbs file or view the macro in the Excel file.

The macro that I created is called **m_format_table**. It formats our summary table, so that it has the standard REPS fonts and coloring. Feel free to run the code below to try it out. Your outputs will be saved here: [R:/!Projects/TRAIN -- REPS Training and PD/R/Lesson 05 Excel Macros/Sample Outputs](R:/!Projects/TRAIN -- REPS Training and PD/R/Lesson 05 Excel Macros/Sample Outputs)

This next block of code will save our summary table into the Excel file (.xlsm format).

```{r Export table, eval=FALSE}
#Name file and path
today_date <- format.Date(Sys.Date(),"%Y%m%d")
last_name <- dlgInput("Enter your last name")$res
first_initial <- dlgInput("Enter your first initial")$res
user_name <- paste0(last_name, first_initial)
save_directory <- "R:/!Projects/TRAIN -- REPS Training and PD/R/Lesson 05 Excel Macros/Sample Outputs/"
new_excel_name <- paste0(save_directory,"sample_summary_table_", user_name, "_", today_date, ".xlsm")

#Copy template and save as new file
file.copy(from="R:/!Projects/TRAIN -- REPS Training and PD/R/Lesson 05 Excel Macros/Sample VBA Macro/excel_with_macro.xlsm", to=new_excel_name,overwrite= TRUE)
write.xlsx2(as.data.frame(summary_table), new_excel_name, sheetName = "summary.table", row.names=FALSE, append=TRUE)
dlgMessage(paste0("Your file has been saved here: ", new_excel_name))
```
Before continuing, I recommend opening the Excel file and viewing the output. You will notice that the output is plain. It's just a raw data table thrown into an Excel file.

This next block of code will run the Excel macro, which will format our data table. This code takes advantage of the *RDCOMClient* package in R. It creates a server that allows us to make changes to an Excel file from R. There is a LOT of back end coding that we do not have to do as a result. As you will notice in this next code block, seven lines of code will allow us to run the macro that we have created in Excel from our R script.
```{r VBA macro, eval=FALSE, message=FALSE}
#Create COM server
xlApp <- COMCreate("Excel.Application")

#Open excel worksheet
xlWbk <- xlApp$Workbooks()$Open(new_excel_name)
xlSheet <- xlWbk$Sheets("summary.table")$Activate()

#Run macro
xlApp$Run("m_format_table")

#Close and save
xlWbk$Close(TRUE)
xlApp$Quit()
rm(xlWbk, xlApp)
```

#Another example...
What is the utility of Excel macros? In this example, we run the exact same code as above, but with one caveat. This time, the shape of the data is different. We have another cohort, which means our summary table will be longer. Run the first block of code and then view the summary_table_02 result. If we were to put the data table into a pre-existing template, that means having to re-do tedious formatting work. What if we can automate this processe with Excel macros? 
```{r Example Two: Summary Table, eval=FALSE}

sample_data_02 <- read.xlsx("R:\\!Projects\\TRAIN -- REPS Training and PD\\R\\Lesson 05 Excel Macros\\Sample Data\\r_train_excel_macro_sample_data_20170720.xlsx", sheetIndex = 2)

summary_table_02 <- sample_data_02 %>% 
  mutate(Cohort = paste0(Semester," ",Year),
         grad_dummy = if_else(GRAD=="Yes", 1, 0)) %>% 
  group_by(Cohort) %>% 
  summarise(count = n(),
            Average.GPA = mean(mean_GPA),
            Grad.Rate = sum(grad_dummy)/count) %>% 
  ungroup()

summary_all_02 <- summary_table_02 %>% 
  summarise(Cohort = "All cohorts", 
            count = sum(count),
            Average.GPA = mean(Average.GPA),
            Grad.Rate = sum(Grad.Rate)/count)

summary_table_02 <- summary_table_02 %>% 
  bind_rows(summary_all_02)
```
Let's run the Excel macro and see what happens...
```{r Example Two: Running Excel macro, eval=FALSE}
write.xlsx2(as.data.frame(summary_table_02), new_excel_name, sheetName = "summary.table.2", row.names=FALSE, append=TRUE)

xlApp <- COMCreate("Excel.Application")
xlWbk <- xlApp$Workbooks()$Open(new_excel_name)
xlSheet <- xlWbk$Sheets("summary.table.2")$Activate()
xlApp$Run("m_format_table")
xlWbk$Close(TRUE)
xlApp$Quit()
rm(xlWbk, xlApp)
```

